Future Improvement

### 2020-09-14 tensor_regress/oldcode/new_manuscript/tensor_new.tex
clean

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url}

\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{algorithm}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor={black},citecolor={blue},urlcolor={blue}}

\newcommand{\blind}{1}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{-.3in}%
\addtolength{\topmargin}{-.8in}%

@@

\usepackage{appendix}
\usepackage{wrapfig}
\mathtoolsset{showonlyrefs}

@@

\def\ci{\perp\!\!\!\perp}

\def\fixme#1#2{\textbf{\color{red}[FIXME (#1): #2]}}
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Supervised Tensor Decomposition with Interactive Side Information}
  %\author{Jiaxin Hu, Zhuoyan Xu, and Miaoyan Wang\\
   % Department of Statistics, University of Wisconsin-Madison}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Supervised Tensor Decomposition with Interactive Side Information}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Higher-order tensors have received increased attention across science and engineering. While most tensor decomposition methods are developed for a single tensor observation, scientific studies often collect side information, in the form of node features and interactions thereof, together with the tensor data. Such data problems are common in neuroimaging, network analysis, and spatial-temporal modeling. Identifying the relationship between a high-dimensional tensor and side information is important yet challenging. Here, we develop a tensor decomposition method that incorporates multiple side information as interactive features. Unlike unsupervised tensor decomposition, our supervised decomposition captures the effective dimension reduction of the data tensor confined to feature space on each mode. An efficient alternating optimization algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. We apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We identify the key global connectivity pattern and pinpoints the local regions that are associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, the package and data used are available at~\url{https://CRAN.R-project.org/package=tensorregress}.

\end{abstract}

\noindent%
{\it Keywords:} Applications and case studies, Tensor data analysis, Supervised dimension reduction, Exponential family distribution, Generalized multilinear model
\vfill

\newpage
\spacingset{1.5} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Multi-dimensional arrays, known as tensors, are often collected with side information on multiple modes in modern scientific and engineering studies. A popular example is in neuroimaging~\citep{sun2017store,zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\citep{baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent the friendships. Side information such as people’s demographic information and friendship types are often available. In both examples, it is of keen scientific interest to identify the variation in the tensor data (e.g., brain connectivities, social community patterns) that is affected by available features. These seemingly different scenarios pose a common yet challenging problem for tensor data modeling.

In addition to the aforementioned challenges, many instances of tensor datasets consist of non-Gaussian measurements. Examples include the political interaction dataset \citep{hu2015scalable} which measures action counts between countries under various relationships, and the brain connectivity network dataset \citep{wang2019common} which is a collection of binary adjacency matrices. Classical tensor decomposition methods are based on minimizing the Frobenious norm of the reconstruction error, leading to suboptimal predictions for binary- or count-valued response variables. A number of supervised tensor methods have been proposed \citep{narita2012tensor, zhao2012higher, yu2016learning} to address the tensor regression problem in various forms (e.g.\ scalar-to-tensor regression, tensor-response regression). These methods often assume Gaussian distribution for the tensor entries, or impose random designs for the feature matrices, both of which are less suitable for applications of our interest. The gap in theory and practice suggests a great opportunity for improved modeling paradigms that better capture the complexity in tensor data with side information.

\begin{figure*}[t]
\begin{center}
\includegraphics[width=16cm]{new_introfigure1.pdf}
\end{center}
\caption{Examples of supervised tensor decomposition with interactive side information. (a) Network population model. (b) Spatial-temporal growth model. }\label{fig:intro1}
\end{figure*}

We present a general model and associated method for decomposing a data tensor with exponential family entries and interactive side information. 
We formulate the learning task as a structured regression problem, with tensor observation serving as the response, and the multiple side information as interactive features. Figure~\ref{fig:intro1}b illustrates our model in the special case of order-3 tensors. A low-rank structure is imposed to the conditional mean tensor, where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

In addition, we use tools from generalized linear model (GLM) to allow heteroscedacity due to the mean-variance relationship in the non-Gaussian data. This flexibility is important in practice. In classical GLM, the sample size and feature dimension are well defined; however, in the tensor data analysis, we observe only one single realization of tensor response but up to $K$ interactive feature matrices, where $K$ is the number of tensor modes. Both the sample size and feature dimension grow exponentially in $K$. We use a low-rank constraint in the conditional mean tensor (with a suitable transformation) to mitigate the curse of high dimensionality. The statistical convergence of our estimator is established, and we quantify the gain in prediction through several case studies.

Our work is closely related to but also clearly distinctive from several lines of previous work. The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of a data tensor. In contrast, our model is a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and multiple features. The low-rank factorization is determined jointly by the tensor data and feature matrices.

The second line of work studies the scalar-to-tensor regression, in which the response is a scalar and the predictor is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to this line because we treat the data tensor as random variables. The difference in model settings yields different focuses and interpretations. Consider the neuroimaging analysis as an example. The scalar-to-tensor regression aims to predict the people's disease status based on brain connectivity network, while tensor-to-scalar model focuses on understanding the variation in the brain connectivity attributable to the individual features such as age, gender, and disease status. The scalar-to-tensor model is insufficient for tensor dimension reduction because it lacks the modeling of stochastic noise in the data tensor.

The third line of work studies the tensor-on-tensor regression, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, gahrooei2020multiple}. Our model shares a common ground but provides a more efficient context-specific solutions than earlier approaches. As we show in Section~\ref{subsec:statprob}, the supervised tensor decomposition 
has an interesting interpretation as a special tensor-on-tensor regression. Nevertheless, our work improves from previous work in scope and applicability. Previous methods~\citep{gahrooei2020multiple,lock2018tensor} mainly focus on Gaussian tensors. The Frobenius norm used in the objective function is statistical suboptimal for general exponential family tensors. Maximum likelihood estimator (MLE) is studied in~\cite{raskutti2015convex} and a convex relaxation algorithm is proposed to solve for low-rank tensor coefficients. However, in the tensor case, convex MLE suffers from both computational intractability and statistical suboptimality. We advocate a non-convex approach and provide strong evidence for its success in our setting. Most previous tensor regression focuses on prediction~\citep{lock2018tensor,raskutti2015convex,gahrooei2020multiple}, and we go step further by finding the sufficient dimension reduction~\citep{adragni2009sufficient}, $\text{Span}(\mM_k)$, that facilitates the identification of effective features in prediction (see Figure~\ref{fig:intro1}b). The latter approach greatly improves the \emph{interpretability} in prediction. In this regards, our method opens up new opportunities for tensor data analysis in a wider range of possible applications.

\fixme{Miaoyan}{One more paragraph. Current literature on tensor decomposition with auxiliary side information...}

The remainder of this paper is organized as follows. Section~\ref{sec:pre} introduces tensor notation and preliminaries. Section~\ref{sec:model} presents the main model and three motivating examples for supervised tensor decomposition. Section~\ref{sec:est} describes a rank-constrained likelihood-based estimation and associated alternating optimization algorithm. In Section~\ref{sec:simulation}, we present numerical experiments and assess the performance in comparison to alternative methods. In Section~\ref{sec:data}, we apply the method to diffusion tensor imaging data from human connectome project and multi-relational social network data. We conclude in Section~\ref{sec:con} with discussions about our findings and avenues of future work.

\section{Preliminaries}\label{sec:pre}

We introduce the basic tensor properties used in the paper; more details on tensor notation can be found in~\citet{kolda2009tensor}. We use $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$\fixme{Miaoyan}{perhaps use $y_\omega$} to denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor, where $K$ corresponds to the number of modes of $\tY$ and is called the order. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as
\begin{equation}
\tY \times_1\mX_1\ldots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\ldots x_{j_K,i_K}^{(K)}},
\end{equation}
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use the shorthand $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as 
\[
\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}.
\] 
The Frobenius norm and maximum norm of $\tY$ are defined as 
\[
\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}\quad \text{and}\quad \mnormSize{}{\tY}=\max_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}.
\]
 A higher-order tensor can be reshaped into a lower-order object. We use $\text{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ the operation that reshapes the tensor along mode-$k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The multilinear rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\text{Unfold}_k(\tY)$, $k=1,\ldots,K$. Given a matrix $\mM\in\mathbb{R}^{d\times r}$, we use $\text{Span}(\mM)$ to denote the space spanned by columns of $\mM$. We use lower-case letters (e.g.,\ $a,\ b,\ c$) for scalars and vectors, upper-case boldface letters (e.g.,\ $\mA,\ \mB,\ \mC$) for matrices, and calligraphy letters (e.g.,\ $\tA,\ \tB,\  \tC$) for tensors of order three or greater. We let $\mI_d$ denote the $d \times d$ identity matrix and $[d]$ denote the $d$-set $\{1,\ldots,d\}$. For ease of notation, we allow the basic arithmetic operators (e.g.,\ $+$, $-$, $\geq $) and univariate 
functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner.

\section{Motivation and model}\label{sec:model}
\subsection{General framework for supervised tensor decomposition}
We begin with a general framework for supervised tensor decomposition and then discuss its implication in three concrete examples. Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We propose a multilinear conditional mean model between the data tensor and feature matrices. Assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\quad \text{with}\quad \Theta =\tB\times\{\mX_1,\ \ldots,\ \mX_K\} ,
\end{align}
where $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the unknown parameter tensor, $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$, and $\times$ denotes the tensor-by-matrix product. The choice of link function is based on the assumed distribution family of tensor entries. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution. In general, dispersion parameters can also be included in the model. Because our main focus is the tensor decomposition under the mean model, we omit the dispersion parameter in this section for ease of presentation.

In classical tensor decomposition, tensor factorization is performed on either data tensor $\tY$ or mean tensor $\mathbb{E}(\tY)$. In the context of supervised tensor decomposition, we propose to factorize the latent parameter tensor $\tB$,
\begin{equation}\label{eq:factor}
\tB=\tC\times\{\mM_1,\ \ldots,\ \mM_K\},
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, and $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns, where $r_k\leq p_k$ for all $k\in[K]$. By the definition of multilinear rank, model equations~\eqref{eq:tensormodel} and~\eqref{eq:factor} imply the low-rankness $\mr=(r_1,\ldots,r_K)$ of the conditional mean tensor under the link function. We now reach our final model for supervised tensor decomposition,
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1,\ \ldots,\ \mX_K\mM_K\}\right),\notag\\
\text{where} \ &\ \mM_k\mM^T_k=\mI_{p_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\ \text{for all }k=1,\ldots,K,
\end{align}
where the parameters of interest are $\mM_k$ and $\tC$. Note that model~\eqref{eq:decomp} assumes a fixed, known rank $\mr=(r_1,\ldots,r_K)$; the adaptation to unknown rank will be addressed in Section~\ref{sec:tuning}. Figure~\ref{fig:intro1}b provides a schematic illustration of our model. The features $\mX_k$ affect the distribution of tensor entries in $\tY$ through the form $\mX_k\mM_k$, which are $r_k$ linear combinations of features on mode $k$. We call $\mX_k\mM_k$ the ``supervised tensor factors'' or ``sufficient features''~\citep{adragni2009sufficient}, and call $\mM_k$ the ``dimension reduction matrix.'' The core tensor $\tC$ collects the interaction effects between sufficient features across $K$ modes. Our goal is to find $\mM_k$ and the corresponding $\tC$. Note that $\mM_k$ and $\tC$ are identifiable only up to orthonormal transformations.

\subsection{Three examples}
We give three concrete examples of supervised tensor decomposition model~\eqref{eq:decomp} that arises in practice. 
\begin{example}[Spatio-temporal growth model]
The growth curve model~\citep{gabriel1998generalised,potthoff1964generalized,srivastava2008models} was originally proposed as an example of bilinear model for matrix data, and we adopt its higher-order extension here. Let $\tY=\entry{y_{ijk}}\in\mathbb{R}^{d \times m\times n}$ denote the pH measurements of $d$ lakes at $m$ levels of depth and for $n$ time points. Suppose the sampled lakes belong to $q$ types, with $p$ lakes in each type. Let $\{\ell_j\}_{j\in[m]}$ denote the sampled depth levels and $\{t_k\}_{k\in[n]}$ the time points. Assume that the expected pH trend in depth is a polynomial of order at most $r$ and that the expected trend in time is a polynomial of order $s$. Then, the conditional mean model for the spatio-temporal growth model can be represented as
\begin{equation}\label{eq:time}
\mathbb{E}(\tY|\mX_1,\mX_2,\mX_3)=\tC\times\{\mX_1\mM_1,\ \mX_2\mM_2,\ \mX_3\mM_3\},
\end{equation}
where $\mX_1=\text{blockdiag}\{\mathbf{1}_p,\ldots,\mathbf{1}_p\}\in \{0,1\}^{d\times q}$ is the design matrix for lake types, and
\[
\mX_2=
\begin{pmatrix}
1 & \ell_1&\cdots &\ell^{r}_1\\
1 & \ell_2&\cdots &\ell^{r}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&\ell_{m}&\cdots&\ell^{r}_{m}
\end{pmatrix},\quad
\mX_3=
\begin{pmatrix}
1 & t_1&\cdots &t^{s}_1\\
1 & t_2&\cdots &t^{s}_2\\
\vdots &\vdots&\ddots&\vdots\\
1&t_{n}&\cdots&t^{s}_{n}
\end{pmatrix}
\]
are the design matrices for spatial and temporal effects, respectively, $\tC\in\mathbb{R}^{r_1\times r_2\times r_3}$ is the unknown core tensor, and $\mM_k$ are unknown dimension reduction matrices of coherent sizes. The factors $\mX_k\mM_k$ are sufficient features for the mean model~\eqref{eq:time}. The spatial-temporal model is a special case of our supervised tensor decomposition model, with features available on each of the three modes.
\end{example}

\begin{example}[Network population model]\label{example:brain}
Network response model~\citep{rabusseau2016low, zhang2018network} is recently developed for neuroimanig analysis. The goal is to study the relationship between brain network connectivity pattern and features of individuals. Suppose we have a sample of $n$ observations, $\{(\mY_i, \mx_i)\colon i=1,\ldots,n\}$, where for each individual $i\in[n]$, $\mY_i\in\{0,1\}^{d\times d}$ is the symmetric adjacency matrix whose entries indicate presences/absences of connectivities between $d$ brain nodes, and $\mx_i\in\mathbb{R}^p$ is the individual's feature such as age, gender, cognition score, etc. The network-response model  has the conditional mean
\begin{equation}\label{eq:network}
\text{logit}(\mathbb{E}(\mY_i|\mx_i))=\tB\times_3\mx_i, \quad \text{for }i=1,\ldots,n,
\end{equation}
where $\tB\in \mathbb{R}^{d\times d\times p}$ is a coefficient tensor of rank $r(\tB)=(r,r,r')$.

The model~\eqref{eq:network} is a special case of our supervised tensor decomposition, with feature matrix on the last mode of the tensor. Specifically, we stack the network observations $\{\mY_i\}$ together and obtain an order-3 response tensor $\tY\in\{0,1\}^{d\times d\times n}$. Define a feature matrix $\mX=[\mx_1,\ldots,\mx_n]^T\in\mathbb{R}^{n\times p}$. Then, the model~\eqref{eq:network} has the equivalent representation of supervised tensor decomposition,
\[
\text{logit}(\mathbb{E}(\tY|\mX))=\tC\times\{\mM,\ \mM,\ \mX\mM'\},
\]
where $\tC\in\mathbb{R}^{r\times r\times r'}$ is the core tensor, $\mM\in\mathbb{R}^{d\times r}$ is the dimension reduction matrix at the first two modes, and $\mM'\in\mathbb{R}^{p\times r'}$ is for the last mode.  \end{example}

>>>>

\documentclass[11pt]{article}

\usepackage{fancybox}
\usepackage{color}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{comment}
\usepackage{amsmath,amssymb,amsthm,bm,mathtools}
\usepackage{natbib}
\usepackage{dsfont,multirow,hyperref,setspace,enumerate}
\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={red}}
\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{makecell}

\setlength\parindent{0pt}
\usepackage[parfill]{parskip}

@@

\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\usepackage{wrapfig}
\mathtoolsset{showonlyrefs}

%\linespread{1.5}

\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{booktabs}       % professional-quality tables
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{subfig}
%\usepackage{algorithmic}
\usepackage{appendix}

@@

%\externaldocument{tensor_regression_supp}

 %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%% %%%
\begin{document}

\begin{center}

{\Large \vspace*{.3cm}
Exponential Family Tensor Regression with Covariates on Multiple Modes}

\vspace*{1cm}
Jiaxin Hu, Zhuoyan Xu, and Miaoyan Wang\footnote{Miaoyan Wang is Assistant Professor, Department of Statistics, University of Wisconsin-Madison, Madison, WI 53706, E-mail: miaoyan.wang@wisc.edu. Jiaxin Hu is a Ph.D. student in Statistics, E-mail: jhu267@wisc.edu. Zhuoyan Xu is a Ph.D. student in Statistics, E-mail: zxu444@wisc.edu. }

\vspace*{5mm}
Department of Statistics, University of Wisconsin-Madison
\end{center}

\vspace*{.5cm}

\begin{abstract}
We consider the problem of tensor-response regression given covariates on multiple modes. Such data problems arise frequently in applications such as neuroimaging, network analysis, and spatial-temporal modeling. We propose a new family of tensor response regression models that incorporate covariates, and establish the theoretical accuracy guarantees. Unlike earlier methods, our estimation allows high-dimensionality in both the tensor response and the covariate matrices on multiple modes. An efficient alternating updating algorithm is further developed. Our proposal handles a broad range of data types, including continuous, count, and binary observations. Through simulation and applications to two real datasets, we demonstrate the outperformance of our approach over the state-of-art. \end{abstract}

{\bf Keywords:} Tensor-response models, Multiple covariates, Dimension reduction, Reduced-rank regression, Generalized linear models

\section{Introduction}

Nowadays, multi-dimensional arrays, a.k.a.\ tensors, are collected with additional covariates on multiple modes in many modern scientific and engineering studies. One example is neuroimaging~\citep{sun2017store,zhou2013tensor}. The brain connectivity networks are collected from a sample of individuals, accompanied by individual characteristics such as age, gender, and diseases status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\citep{baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent the friendships. Other covariates such as people’s demographic information and the closeness of a friendship are often available. In both examples, it is of keen scientific interest to identify the variation in the response (e.g., brain connectivity, social community patterns) that may be affected by additional covariates. These seemingly different problems can be formulated as a regression problem in the tensor sense.

Emerging instances of the non-Gaussian datasets also require the effective tools applicable to a wide range of data, such as continuous, binary and count data. Examples include the Political Social science data \citep{hu2015scalable} which is a count dataset, and the brain connectivity network data \citep{wang2019common} which is usually encoded as binary datasets. Easily treating these datasets as Gaussian-valued data will yield unstable estimations and lead the predictions fall outside the valid range. A number of previous methods have been proposed \citep{yu2016learning, llosa2018tensor, raskutti2015convex, zhao2012higher} to address the tensor regression problem in various forms (e.g.\ scalar-to-tensor regression, tensor-response regression) from distinct perspectives. However, these methods assume the Gaussianity in the model, treating the input as Gaussian valued entries. Novel methods for exponential family tensor regression are necessary this time.

\begin{figure*}[t]
\begin{center}
\includegraphics[width=16cm]{demo.pdf}
\end{center}
\caption{Examples of tensor response regression model with covariates on multiple modes. (a) Network population model. (b) Spatial-temporal growth model. }\label{fig:intro1}
\end{figure*}

{\bf Related work.} Our work is closely related to but also clearly distinctive from several lines of previous work.

The first line is a class of \textit{unsupervised} tensor decomposition such as Tucker and CP decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor, hong2020generalized} that aims to find the best low-rank representation of data tensor. In contrast, our model can be viewed a \textit{supervised} tensor learning, which aims to identify the association between a data tensor and covariates. The low-rank structure is determined jointly by the tensor response and matrix covariates.

The second line of work studies the scalar-to-tensor regression models, in which the response is a scalar and the \textit{predictor} is a tensor~\citep{zhou2013tensor,chen2019non}. Our proposal is orthogonal to theirs because we treat the tensor as a \textit{response}. The distinct model settings yield distinct focuses and interpretations. Take neuroimaging analysis as an example. The scalar-to-tensor regression aims to study the changes of the prediction outcomes when the brain connectivity network varies, while tensor-response model focuses on understanding the variation in the brain connectivity that may attribute to the individual features such as age, gender and disease status. Since scalar-to-tensor models do not utilize all the information in the response, they are not optimal when there exist correlations among response variables. In contrast, our model does not assume the independence among response entries, which allow the model to capture more structural information.

The third line of work studies the tensor-on-tensor regression models, in which both the response and predictor are tensors~\citep{raskutti2015convex, lock2018tensor, llosa2018tensor, gahrooei2020multiple}. Though our model actually tackles the similar problem as tensor-on-tensor regression, our work differs with previous work in several ways. First, previous methods mainly develop the model with Gaussian error. These methods may not be applicable for other exponential family data, and their theoretical results may not be extended to generalized case trivially. Second, our study differs from previous work in the scope of results. \citep{gahrooei2020multiple} verifies the performance of estimation mainly via simulations, while our study provides a theoretical evidence on estimation. \citep{lock2018tensor, llosa2018tensor} perform solid inferences on parameter estimations, however, they do not assess the distance between estimation and true parameters and the prediction error, which are studied in our paper. Risk bounds are well studied in~\citep{raskutti2015convex} and our convergence rate coincides with their results in the order-3 low-rank case. Whereas, the convexity assumption of the regularization in their model yields a convex optimization problem. In contrast,  our model tackles a non-convex optimization problem, and employee distinct techniques.

{\bf Our contribution.} In this article, we present a general model and the associated theory for exponential family tensor regression. Exponential family tensor observation serves as a response, and the node features and/or the interactions forms the matrix predictor. Figure~\ref{fig:intro1}b illustrates the general set-up we consider. Due to the generalized framework, our models allows heteroscedasticity by relating the variance of tensor entry to its mean. This ﬂexibility is particularly important in practice.  The regression approach allows the identiﬁcation of variation in the data tensor that is explained by the covariates. We utilize a low-rank constraint in the regression coeﬃcient to encourage the sharing among tensor entries. The statistical convergence of our estimator is established, and we quantify the gain in predictive power.

\iffalse
Many contemporary scientific and engineering studies collect multi-way array data, a.k.a.\ tensors, accompanied by additional covariates. One example is neuroimaging  analysis~\citep{sun2017store,zhou2013tensor}, in which the brain connectivity networks are collected from a sample of individuals. Researchers are often interested in identifying connection edges that are affected by individual characteristics such as age, gender, and disease status (see Figure~\ref{fig:intro1}a). Another example is in the field of network analysis~\citep{wu2017generalized,baldin2018optimal,hoff2005bilinear}. A typical social network consists of nodes that represent people and edges that represent friendships. In addition, features on nodes and edges are often available, such as people's personality and demographic location. It is of keen scientific interest to identify the variation in the connection patterns (e.g., transitivity, community) that can be attributable to the node features.

This paper presents a general treatment to these seemingly different problems. We formulate the learning task as a regression problem, with tensor observation serving as a response, and the node features and/or their interactions forming the predictor. Figure~\ref{fig:intro1}b illustrates the general set-up we consider. The regression approach allows the identification of variation in the data tensor that is explained by the covariates. In contrast to earlier work~\citep{zhang2018network,wu2017generalized}, our method allows the contribution of covariates from multiple modes, whenever available. We utilize a low-rank constraint in the regression coefficient to encourage the sharing among tensor entries. The statistical convergence of our estimator is established, and we quantify the gain in accuracy compared to classical multivariate regression approach.

{\bf Related work.} Our work is closely related to but also clearly distinctive from several lines of previous work. The first is a class of \emph{unsupervised} tensor decomposition~\citep{de2000multilinear, kolda2009tensor,zhang2018tensor} that aims to find a low-rank representation of a data tensor. In contrast, our model can be viewed a \emph{supervised} tensor learning, which aims to identify the association between a data tensor and covariates. The second related line~\citep{zhou2013tensor,chen2019non} tackles tensor regression where the response is a scalar and the \emph{predictor} is a tensor. Our proposal is orthogonal to theirs because we treat the tensor as a \emph{response}. The tensor-response model is appealing for high-dimensional analysis when both the response and the covariate dimensions grow. The last line of work studies the network-response model~\citep{rabusseau2016low, li2017parsimonious}. The earlier development of this model focuses mostly on binary data in the presence of dyadic covariates~\citep{hoff2005bilinear,wu2017generalized}. We will demonstrate the enhanced accuracy as the order of data grows, and establish the general theory for exponential family which is arguably better suited to various data types. 
\fi

\vspace{-.1cm}
\section{Preliminaries}
\vspace{-.1cm}
We begin by reviewing the basic properties about tensors~\citep{kolda2009tensor}. We use $\tY=\entry{y_{i_1,\ldots,i_K}}\in \mathbb{R}^{d_1\times \cdots\times d_K}$ to denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as
\begin{equation}
\tY \times_1\mX_1\ldots \times_K \mX_K=\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\ldots x_{j_K,i_K}^{(K)}},
\end{equation}
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. For ease of presentation, we use shorthand notion $\tY\times\{\mX_1,\ldots,\mX_K\}$ to denote the tensor-by-matrix product. For any two tensors $\tY=\entry{y_{i_1,\ldots,i_K}}$, $\tY'=\entry{y'_{i_1,\ldots,i_K}}$ of identical order and dimensions, their inner product is defined as $\langle \tY, \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}$. The Frobenius norm of tensor $\tY$ is defined as $\FnormSize{}{\tY}=\langle \tY,\ \tY \rangle^{1/2}$. A higher-order tensor can be reshaped into a lower-order object~\citep{wang2017operator}. We use $\text{vec}(\cdot)$ to denote the operation that reshapes the tensor into a vector, and $\text{Unfold}_k(\cdot)$ the operation that reshapes the tensor along mode-$k$ into a matrix of size $d_k$-by-$\prod_{i\neq k}d_i$. The Tucker rank of an order-$K$ tensor $\tY$ is defined as a length-$K$ vector $\mr=(r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\text{Unfold}_k(\tY)$, $k=1,\ldots,K$. We use lower-case letters (e.g.,\ $a,b,c$) for scalars/vectors, upper-case boldface letters (e.g.,\ $\mA,\mB,\mC,$) for matrices, and calligraphy letters (e.g.,\ $\tA, \tB, \tC,$) for tensors of order three or greater. We let $\mI_d$ denote the $d \times d$ identity matrix, $[d]$ denote the $d$-set $\{1,\ldots,d\}$, and allow an $\mathbb{R}\to \mathbb{R}$ function to be applied to tensors in an element-wise manner.

\section{Motivation and model}\label{sec:model}
Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose we observe covariates on some of the $K$ modes. Let $\mX_k\in\mathbb{R}^{d_k\times p_k}$ denote the available covariates on the mode $k$, where $p_k\leq d_k$. We propose a multilinear structure on the conditional expectation of the tensor. Specifically, 
\begin{align}\label{eq:tensormodel}
&\mathbb{E}(\tY|\mX_1,\ldots,\mX_K)=f(\Theta),\ \text{with}\\
&\Theta =\tB\times\{\mX_1,\ldots,\mX_K\} ,
\end{align}
where  $\Theta\in\mathbb{R}^{d_1\times \cdots\times d_K}$ is the multilinear predictor, $\tB\in\mathbb{R}^{p_1\times \cdots \times p_K}$ is the parameter tensor of interest, $f(\cdot)$ applied to $\Theta$ entrywise is a known link function in generalized linear model, and $\times$ denotes the tensor Tucker product. The choice of link function depends on the distribution of the response data. Some common choices are identity link for Gaussian tensor, logistic link for binary tensor, and $\exp(\cdot)$ link for Poisson tensor (see Table~\ref{table:link}).
